Abstractions for Multi-Sorted Substitutions

### 2024-03-13 paper/paper.tex
Update.

  A compose kit \ACode{\ACon{CKit}~K₁~K₂~K₁⊔K₂}
  describes the operations necessary for definining the composition of a
  map from \ACode{K₁} with a map from \ACode{K₂} that results in a map from
  \ACode{K₁⊔K₂}:
  \LibCode\KCKit
  \noindent The third parameter \ACode{K₁⊔K₂} can be seen as a functional
  dependency\cite{DBLP:conf/esop/Jones00} and is determined
  by the choice of \ACode{K₁} and \ACode{K₂}.
  The fields of a compose kit have the following meaning:
  \begin{itemize}
  \item
    The \ACode{\AField{\_\&/⋯\_}} operation takes a variable or term
    \ACode{x/t} (according to \ACode{K₁}) and a renaming or substitution
    \ACode{ϕ} (according to \ACode{K₂}) and applies \ACode{ϕ} to \ACode{x/t} resulting in
    a variable or term (according to \ACode{K₁⊔K₂}).
    From this operation we derive map composition \ACode{\ADef{\_·ₖ\_}} as
    shown in the previous subsection.
    % TODO: use the notation def to the first time we use it?
    We use the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}, if we need to
    make it explicit that we are using the composition of some
    specific \ACode{\ACon{CKit}} \ACode{C}.
  \item 
    The \ACode{\AField{\&/⋯-⋯}} lemma describes the behavior of
    \ACode{\AField{\_\&/⋯\_}} in terms of \ACode{\AField{\_⋯\_}}, allowing
    subsequent lemmas to make use of the lemmas that we have already proved
    for \ACode{\AField{\_⋯\_}}.
  \item 
    The \ACode{\AField{\&/⋯-wk-↑}} lemma states that applying a map and then
    weakening is the same as weakening first and then lifting the map over
    the variable introduced by the weakening.
    From this lemma, we can derive that lifting distributes over composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}

@@

  %   % The \ACode{\AField{`/id}} on the left side is from \ACode{K₁⊔K₂}, whereas
  %   % the \ACode{\AField{`/id}} on the right side is from \ACode{K₂}.
  \item
    We show that lifting \ACode{\AField{\_↑\_}} distributes over
    composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}
  To make it easier to explicitly specify a \ACode{\ACon{CKit}},
  we define the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}
  to stand for \ACode{ϕ₁ \ADef{·ₖ} ϕ₂} for some
  \ACode{\ACon{CKit}} \ACode{C}.
  % \LibCode\KCKitNotation

@@

  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor, which covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing \ACode{\ACon{⊢λ}} of the lambda abstraction, which
    weakens the type of the codomain \ACode{t₂}.
    This is necessary, because multi-sorted syntax allows types
    to depend on expressions, so the typing derivation for \ACode{e}
    has to account for a variable, which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-typed due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  The instantiation for System F is straightforward:
  \AppCode\FTypingInst

@@

  We start with \ACode{\ACon{TKit}}s, which abstract over
  variable and term \emph{typing}, and then define a
  \ACode{\ACon{TTraversal}}, which provides
  substitution-preserves-typing for all \ACode{\ACon{TKit}}s.
  \LibCode\KTKit
  The operations are completely analogous to those of a \ACode{\ACon{Kit}}, but talk
  about typing instead of terms.
  Building on the fields of a \ACode{\ACon{TKit}}, we define further operations:
  \begin{itemize}
  \item
    Using the variable/term typing, we can define a renaming/substitution typing:
    \LibCode\KMapTyping
    \ACode{Γ₂ \ADef{∋*/⊢*} ϕ \ADef{∶} Γ₁} states that \ACode{ϕ} is a map that takes
    terms in \ACode{Γ₁} to terms in \ACode{Γ₂}.
  \item 
    Lifting a map preserves its typing:
    \LibCode\KLiftTyping
    % \LibCode\KLiftTypingProof
  \item 
    If a variable/term has a typing, then so does its singleton renaming/substitution:
    \LibCode\KSingleTyping
    % \LibCode\KSingleTypingProof
  \end{itemize}
  To make it easier to explicitly pass typing kits,
  we define the notation \ACode{Γ₂ \ADef{∋*/⊢*[} TK \ADef{]} ϕ \ADef{∶} Γ₁}
  to stand for \ACode{Γ₂ \ADef{∋*/⊢*} ϕ \ADef{∶} Γ₁} given some
  \ACode{\ACon{TKit}} \ACode{TK}.
  % \LibCode\KTypingNotation

>>>>

  A compose kit \ACode{\ACon{CKit}~K₁~K₂~K₁⊔K₂}
  describes the operations necessary for definining the composition of a
  map from \ACode{K₁} with a map from \ACode{K₂} that results in a map from
  \ACode{K₁⊔K₂}:
  \LibCode\KCKit
  \noindent The third parameter \ACode{K₁⊔K₂} can be seen as a functional
  dependency\cite{DBLP:conf/esop/Jones00} and is determined
  by the choice of \ACode{K₁} and \ACode{K₂}.
  The fields of a compose kit have the following meaning:
  \begin{itemize}
  \item
    The \ACode{\AField{\_\&/⋯\_}} operation takes a variable or term
    \ACode{x/t} (according to \ACode{K₁}) and a renaming or substitution
    \ACode{ϕ} (according to \ACode{K₂}) and applies \ACode{ϕ} to \ACode{x/t} resulting in
    a variable or term (according to \ACode{K₁⊔K₂}).
    From this operation we derive map composition \ACode{\ADef{\_·ₖ\_}} as
    shown in the previous subsection.
    % We use the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}, if we need to
    % make it explicit that we are using the composition of some
    % specific \ACode{\ACon{CKit}} \ACode{C}.
  \item 
    The \ACode{\AField{\&/⋯-⋯}} lemma describes the behavior of
    \ACode{\AField{\_\&/⋯\_}} in terms of \ACode{\AField{\_⋯\_}}, allowing
    subsequent lemmas to make use of the lemmas that we have already proved
    for \ACode{\AField{\_⋯\_}}.
  \item 
    The \ACode{\AField{\&/⋯-wk-↑}} lemma states that applying a map and then
    weakening is the same as weakening first and then lifting the map over
    the variable introduced by the weakening.
    From this lemma, we can derive that lifting distributes over composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}

@@

  %   % The \ACode{\AField{`/id}} on the left side is from \ACode{K₁⊔K₂}, whereas
  %   % the \ACode{\AField{`/id}} on the right side is from \ACode{K₂}.
  \item
    We show that lifting \ACode{\AField{\_↑\_}} distributes over
    composition:
    \LibCode\KDistLiftCompose
    % \LibCode\KDistLiftComposeProof
  \end{itemize}
  % To make it easier to explicitly specify a \ACode{\ACon{CKit}},
  % we define the notation \ACode{ϕ₁ \ADef{·[} C \ADef{]} ϕ₂}
  % to stand for \ACode{ϕ₁ \ADef{·ₖ} ϕ₂} for some
  % \ACode{\ACon{CKit}} \ACode{C}.
  % \LibCode\KCKitNotation

@@

  The interesting cases are:
  \begin{itemize}
  \item
    the variable typing constructor, which covers both expression- and
    type-variables, analogously to the variable term constructor;
  \item
    the typing \ACode{\ACon{⊢λ}} of the lambda abstraction, which
    weakens the type of the codomain \ACode{t₂}.
    This is necessary, because multi-sorted syntax allows types
    to depend on expressions, so the typing derivation for \ACode{e}
    has to account for a variable, which is not used by the type.
  \item
    the kinding constructor \ACode{\ACon{⊢τ}} states that all types
    have kind \ACode{\ACon{★}}. This is sufficient for System F as types
    are automatically well-kinded due to intrinsic scoping.
  \end{itemize}
  To teach the framework about typing, we create a structure analogously to
  \ACode{\ACon{Syntax}}:
  \LibCode\KTyping
  The instantiation for System F is straightforward:
  \AppCode\FTypingInst

@@

  We start with \ACode{\ACon{TKit}}s, which abstract over
  variable and term \emph{typing}, and then define a
  \ACode{\ACon{TTraversal}}, which provides
  substitution-preserves-typing for all \ACode{\ACon{TKit}}s.
  \LibCode\KTKit
  The operations are completely analogous to those of a \ACode{\ACon{Kit}}, but talk
  about typing instead of terms.
  Building on the fields of a \ACode{\ACon{TKit}}, we define further operations:
  \begin{itemize}
  \item
    Using the variable/term typing, we can define a renaming/substitution typing:
    \LibCode\KMapTyping
    \ACode{ϕ \ADef{∶} Γ₁ \ADef{⇒ₖ} Γ₂} states that \ACode{ϕ} is a map that takes
    terms in \ACode{Γ₁} to terms in \ACode{Γ₂}.
  \item 
    Lifting a map preserves its typing:
    \LibCode\KLiftTyping
    % \LibCode\KLiftTypingProof
  \item 
    If a variable/term has a typing, then so does its singleton renaming/substitution:
    \LibCode\KSingleTyping
    % \LibCode\KSingleTypingProof
  \end{itemize}
  % To make it easier to explicitly pass typing kits,
  % we define the notation \ACode{ϕ \ADef{∶} Γ₁ \ADef{⇒[} TK \ADef{]} Γ₂}
  % to stand for \ACode{ϕ \ADef{∶} Γ₁ \ADef{⇒ₖ} Γ₂} of some
  % \ACode{\ACon{TKit}} \ACode{TK}.
  % \LibCode\KTypingNotation
